---
title: "Geocentric Models"
subtitle: "Chapter 4 in McElreath (2020)"
author: "Daniela Palleschi"
---

My notes for this chapter are a bit muddled, I did a mix of reading the textbook and watching the YouTube video. Unfortuantely the YouTube video doesn't map onto the chapter very transparently and even has code that isn't in the current textbook version, so some concepts might be repeated/out of order.

## Resources {.unnumbered}

- [YouTube lecture (2023): Geocentric Models](https://www.youtube.com/watch?v=tNOu-SEacNU&list=PLDcUM9US4XdPz-KxHM4XHt7uUVGWWVSus&index=4&t=1208s)
- Solomon Kurz: [Linear models](https://bookdown.org/content/3890/linear-models.html)

## Set-up {.unnumbered}

```{r}
pacman::p_load(
  rethinking,
  tidyverse,
  janitor,
  knitr,
  kableExtra
)
```

```{r}
theme_set(theme_bw())
```


## Permission to be confused {.unnumbered}

From ~ 20 minute mark from the YouTue video

- you don't need to understand everything all at once
  + math, concepts, code, installation problems etc.
- it's important to keep some *flow*
  + keep moving forward despite the resistance of confusion/struggle
  + things shouldn't be too hard, because then you stop moving
  + if it's too easy, you don't feel any resistance
  + feeling confused means you're paying attention :)


## The Geocentric Model {.unnumbered}

- planets have been observed to have a zig-zag orbit from the perspective of Earth
- Geocentric model/Telomeric model
  + used to predict the pattern of planets orbit, with the prior of Earth at the centre of the solar system
- the geocentric model was very accurate
  + however, the orbits of the planets are elliptical and orbit around the *sun*, not the earth
- so models built on inaccurate assumptions can still be very accurate in their predictions
- the orbits of the planets are embedded witin each other (based on their proximity to the sun), so there is a point when two planets' orbits are closer to each other, i.e., when their orbits have them closer together
  + which is how the perception of Mars' zig-zag orbit (and other planets, but it's more stricking for Mars)

- in 1809: Bayesian argument for normal error and least-squares estimation from Gauss

- **Geocentric**: describes association, makes predictions, but mechanisticalistically wrong
  + the problem is not the model itself, rather some external causal model we project onto the model
  + geocentric models are still extremely useful, what's wrong is believing they're *true* reflections of e.g., the solar system
  + there's also nothing wrong with linear regression itself, as long as we don't believe they're an accurate mechanistic model of the system we're studying
- **Gaussian**: abstracts from generative error model, replaces with normal distribution, mechanistically silent
  + usually a pretty good approximation of associations
  + we will be using gaussian error models

- useful when handled *with care*
  + many special cases: ANOVA, ANCOVA, t-test, etc.

## Why normal distributions are normal

### Normal by multiplication

### Normal by log-multiplication

### Using Gaussian distributions

## A language for describing models

## A Gaussian model of height

```{r}
# load Howell1 data from rethinking package
data(Howell1) 
df_weight <- Howell1
```

```{r}
head(df_weight)
```

```{r}
summary(df_weight)
```

- age ranges from babies to seniors
- sex as binary
- height in cm
- weight in kg

`precis()` function from rethinking package:

```{r}
precis(df_weight) |> 
  as_tibble() |> 
  kable(digits = 1) |> 
  kable_styling()
```

To define heights as normally distributed, we write \ref{eq-height-gauss}, where $i$ stands for index. \ref{eq-height-gauss} is saying that all the golem (i.e., our model) knows about each observed height measurement is defined by the same normal distribution with mean $\mu$ and standard deviation $\sigma$.

\begin{align}
h_i \sim Normal(\mu, \sigma) \label{eq-height-gauss} \\
\mu \sim Normal(178, 20) \label{eq-height-prior-mu} \\
\sigma \sim Uniform(0, 50) \label{eq-height-prior-sigma}
\end{align}

\ref{eq-height-prior-mu} represents our *prior* for $\mu$, where our assumed mean is 178 cm (because Richard McElreath is 178cm tall), with a 95\% probability between +/- 40cm (because if the SD is 20cm, 20 x 1.96 is approximately 20 x 2).

```{r}
# R chunk 4.10

# subset data for adults only
d2 <- df_weight |> filter(age > 18)
```


We can plot our priors:

```{r}
# R code chunk 4.12
curve(dnorm(x, 178, 20), from = 100, to = 250)
```


Or, with ggplot (from [Solomon Kurz](https://bookdown.org/content/3890/linear-models.html#a-gaussian-model-of-height)):

```{r}
ggplot(data = tibble(x = seq(from = 100, to = 250, by = .1)), 
       aes(x = x, y = dnorm(x, mean = 178, sd = 20))) +
  geom_line() +
  ylab("density")
```

We can also view our prior for $\sigma$, which is a flat (uniform) prior:

```{r}
# R code chunk 4.13
curve(dunif(x, 0, 50), from = 10, to = 60)
```

And with ggplot (from [Solomon Kurz](https://bookdown.org/content/3890/linear-models.html#a-gaussian-model-of-height)):

```{r}
tibble(x = seq(from = -10, to = 60, by = .1)) %>%
  
  ggplot(aes(x = x, y = dunif(x, min = 0, max = 50))) +
  geom_line() +
  scale_y_continuous(NULL, breaks = NULL) +
  theme(panel.grid = element_blank())
```



A standard deviation must be positiion, and so we've bound it at zero. Picking the upperbound is another question: in our case a sd of 50cm would imply that 95\% of individual heights lie within 100cm of the average height (again, because SD * 1.96), which is a very large range.

Now it's time for a ~prior predictive~ simulation, for which we need our chosen priors for $h, \mu,$ and  $\sigma$, which imply a joint prior distribution of individual heights.

Let's quickly simiulate heights by sampling from our prior. We can process priors just like posteriors, because every posterior is also potentially a prior for a subsequent analysis.

```{r}
# R code 4.14
sample_mu <- rnorm(1e4, 178, 20)
sample_sigma <- runif(1e4, 0, 50)
prior_h <- rnorm(1e4, sample_mu, sample_sigma)
dens(prior_h)
```

And sampling from both $\mu$ and $\sigma$ priors to get a prior probability of heights:

```{r}
n <- 1e4

set.seed(4)
tibble(sample_mu    = rnorm(n, mean = 178,       sd  = 20),
       sample_sigma = runif(n, min  = 0,         max = 50)) %>% 
  mutate(x = rnorm(n, mean = sample_mu, sd  = sample_sigma)) %>% 
  
  ggplot(aes(x = x)) +
  geom_density(fill = "red", linewidth = 0, alpha = .25) +
  scale_y_continuous(NULL, breaks = NULL) +
  labs(subtitle = expression(Prior~predictive~distribution~"for"~italic(h[i])),
       x = NULL) +
  theme(panel.grid = element_blank())
```





### Finding the posterior distribution with `quap`

```{r}
# R chunk 2.23
d3 <- sample( d2$height , size=20 )
```


```{r}
# R chunk 4.24
mu.list <- seq( from=150, to=170 , length.out=200 ) 
sigma.list <- seq( from=4 , to=20 , length.out=200 ) 
post2 <- expand.grid( mu=mu.list , sigma=sigma.list ) 
post2$LL <- sapply( 1:nrow(post2) , function(i) sum( dnorm( d3 , mean=post2$mu[i] , sd=post2$sigma[i] , log=TRUE ) ) )
post2$prod <- post2$LL + dnorm( post2$mu , 178 , 20 , TRUE ) + dunif( post2$sigma , 0 , 50 , TRUE ) 
post2$prob <- exp( post2$prod - max(post2$prod) )
sample2.rows <- sample( 1:nrow(post2) , size=1e4 , replace=TRUE , prob=post2$prob ) 
sample2.mu <- post2$mu[ sample2.rows ] 
sample2.sigma <- post2$sigma[ sample2.rows ]
plot( sample2.mu , sample2.sigma , cex=0.5 , col=col.alpha(rangi2,0.1) , xlab="mu" , ylab="sigma" , pch=16 )
```

```{r}
# R code 4.26
data(Howell1)
d <- Howell1
d2 <- d[d$age >= 18 , ]
```

And define our model and place it into an `alist`.

```{r}
# R code 4.27
flist <-
  alist(height ~ dnorm(mu , sigma) , 
        mu ~ dnorm(178 , 20) , 
        sigma ~ dunif(0 , 50)
        )
```

Fit a model to the data in `d2`.

```{r}
# R code 4.28
m4.1 <- quap(flist, data = d2)
```

Take a look.

```{r}
# R code 4.29
precis(m4.1)
```

### Simulate weights

From the GitHub script 03_howell etc.

```{r}
data(Howell1)
d <- Howell1

# draw data

blank(bty="n")

col2 <- col.alpha(2,0.8)
plot( d$height , d$weight , col=col2 , lwd=3 , cex=1.2 , xlab="height (cm)" , ylab="weight (kg)" )

d <- d[ d$age>=18 , ]
plot( d$height , d$weight , col=col2 , lwd=3 , cex=1.2 , xlab="height (cm)" , ylab="weight (kg)" )

```

### Describing models

(video c. 31:00)

Conventional statistical model notaton:

1. list the variables in the model
2. define each variable as a deterministic or distributional function of the other variables
  + where does each variable get its value?
  
```{r}
# function to simulate weights of individuals from height 
sim_weight <- function(H,b,sd) {
  U <- rnorm(length(H), 0, sd)
  W <- b*H + U
  return(W)
}
```

- this is a function of unobserved U, which has normal error
- where W is a deterministic function of H and U together

- this can be re-written with standard statistical notation as followed:

\begin{align}
W_i &= \beta H_i + U_i \label{eq-W} \\
U_i &\sim Normal(0,\sigma) \label{eq-U} \\
H_i &\sim Uniform(130,170) \label{eq-H}
\end{align}

Each of these is mirrored in a line of code. Along the left of the equations are the variables, along the right are their definitions. The subscripts indicate individuals (indexing).

- $=$ indicates a **deterministic** relationship
- $\sim$ indicates a **distributional** relationship

Equation \ref{eq-W} is the equation for expected weight. Once we know the values to the right of this equation we can assign values to W. Equation \ref{eq-U} represents Gaussian error with standard deviation sigma.
Equation \ref{eq-H} defines height as uniformly distributed from 130cm to 170cm. These are all defined in the code chunk above, but written in the opposite order. In the code it's written in order of execution, but the statistical conventional notation sometimes has arbitrary order, but conventionally is stated in the opposite direction as the code is written because you start with the most general definition (\ref{eq-W}) which depends on the other variables, and then you see how the variables W depends upon are defined. There's a hierarchy that makes sense.

### Estimator

We want to estimate how the average weight changes with height, i.e., how the average weight changes with height.

\begin{equation}
E(W_i|H_i) = \alpha + \beta H_i
\end{equation}

Where $E(W_i|H_i)$ is the average weight conditional on height, $\alpha$ is the intercept and $\beta$ is our slope. So each height has a different average weight.

### Posterior distribution

Good Bayesians estimate a posterior. The equation below is the same as that for when we tossed the globe, but we just have more unknowns now.

\begin{equation}
Pr(\alpha, \beta, \sigma|H_i, W_i) = \frac
{Pr(W_i|H_i, \alpha, \beta, \sigma)Pr(\alpha, \beta, \sigma)}
{Z}
\end{equation}

- $Pr(\alpha, \beta, \sigma|H_i, W_i)$ is the posterior probability of a specific regression line, which is defined by \alpha, \beta, and \sigma (unobserved variables). H and W are our *observed* data, we don't need a posterior distribution for them, they're observed (the likelihood).
- $Pr(W_i|H_i, \alpha, \beta, \sigma)$: garden of forking data
- $Pr(\alpha, \beta, \sigma)$: our priors
- $Z$: a normalising constant that we don't need to pay too much attention to, it ensures the lefthand side of the equation is a proper probability

W is normally distributed with mean that is a a linear function of H.

These models are typically written as below:

\begin{align}
W_i &= Normal(\mu_i, \sigma) \\
\mu_i &\sim \alpha + \beta H_i
\end{align}

### Grid approximat posterior

All posterior distributions begin with a single observations. There's no minimum number of observations in Bayesian inference.

### Quardratic approximation

- we'll approximate the posterior distribution as a multivariate Gaussian distribution. We can do that because posterior distributions are typically multivariate Gaussian distributions if there are enough observations. A tool that does this for us is called `quap()` in the `statrethinking` package.

The Gaussian distribution is quadratic, and in the log-space the Gaussian distribution is a perfect parabola (i.e., quadratic).

When there are no observations, what does the model believe? These are our priors with no likelihood.

```{r}
#| eval: false
# this wasn't working...
m3.1 <-
  quap(
    alist(
      W ~ dnorm(mu, sigma),
      mu <- a + b*Hm,
      a ~ dnorm(0,10),
      b ~ dunif(0,1),
      sigma ~ dunif(0, 10)
    ),
    data - list(W=W, H=H)
  )
```


## Linear Prediction: Adding a predictor

We'll now add a predictor to our model: weight. So we'll look at how height covaries with weight (our predictor).

A scatter plot of weight by height (with ggplot):

```{r}
ggplot(data = d2, 
       aes(x = weight, y = height)) +
  geom_point(shape = 1, size = 2) +
  theme_bw() +
  theme(panel.grid = element_blank())
```

We now have:


\begin{align}
h_i &\sim Normal(\mu_i, \sigma) \hspace{35pt}likelihood \label{eq-likelihood} \\
\mu_i &\sim \alpha + \beta(x_i - \bar{x}) \hspace{35pt}linear\;model \label{eq-linear} \\
\alpha &\sim Normal(178, 20) \hspace{35pt}\beta \; prior \label{eq-alpha}\\
\beta &\sim Normal(178, 20) \hspace{35pt}\mu \; prior \label{eq-mu}\\
\sigma &\sim Uniform(0, 50) \hspace{35pt}\sigma \; prior \label{eq-sigma}
\end{align}

where the average of $x$ values is $\bar{x}$. Our predictor variablle is $x$, which is a list of measures of the same length as $h$. We now define $\mu$ as a function of the values in $x$ in order to include our predictor, weight, in the model. $\alpha$ is our intercept, $\beta$ is our slope.

## Polynomial regression

## Solomon Kurz

```{r}
library(brms)
```


### Fitting a model with `brm()`

Instead of `statrethinking::map()`

```{r}
d2 <- 
  d %>%
  filter(age >= 18)
```


```{r}
b4.1 <- 
  brm(data = d2, 
      family = gaussian,
      height ~ 1,
      prior = c(prior(normal(178, 20), class = Intercept),
                prior(uniform(0, 50), class = sigma, ub = 50)),
      iter = 2000, warmup = 1000, chains = 4, cores = 4,
      seed = 4,
      file = here::here("fits/b04.01"))
```

```{r}
plot(b4.1) 
```

```{r}
print(b4.1)
```

Or, if we want to use the 89\% intervals used in the textbook:

```{r}
summary(b4.1, prob = .89)
```


Now if we want very narrow priors (sd of 0.1!):

```{r}
b4.2 <- 
  brm(data = d2, family = gaussian,
      height ~ 1,
      prior = c(prior(normal(178, 0.1), class = Intercept),
                prior(uniform(0, 50), class = sigma, ub = 50)),
      iter = 2000, warmup = 1000, chains = 4, cores = 4,
      seed = 4,
      file = here::here("fits/b04.02"))
```

```{r}
plot(b4.2)
```

We see the chains actually look pretty good. Let's look at the fixed effects to easily compare the estimates of the two models:

```{r}
rbind(
  summary(b4.1)$fixed |> 
  as_tibble() |> 
  mutate(model = "b4.1"),
summary(b4.2)$fixed |> 
  as_tibble() |> 
  mutate(model = "b4.2")
)
```

### Sampleing from a `brm()` fit

Instead of using the `vcov()` function from the `statrethinking` package (`brms` has a `vcov()` function but it only gives the first element in the matrix), we can get it after putting the HMC chains in a data frame (which we do with `as_draws_df()`):

```{r}
post <- as_draws_df(b4.1)

head(post)
```

```{r}
select(post, b_Intercept:sigma) %>% 
  cov()
```










## Practice

### E1 {.unnumbered}

> In the model definition below, which line is the likelihood?
\begin{align}
y_i &\sim Normal(\mu,\sigma) \label{ex-e1a} \\
\mu &\sim Normal(0,10) \label{ex-e1b} \\
\sigma &\sim Exponential(1) \label{ex-e1c}
\end{align}

\ref{ex-e1a} is the likelihood (\ref{ex-e1b} is the prior for $\mu$ and \ref{ex-e1c} for $\sigma$)

### E2 {.unnumbered}

> In the model definition just above, how many parameters are in the posterior distribution?

There are two parameters in the posterior distribution: $\mu$ and $\sigma$

### E3 {.unnumbered}

> Using the model definition above, write down the appropriate form of Bayes’ theorem that includes the proper likelihood and priors.

Bayes' theorem is:

\begin{equation}
P(A|B) = \frac{P(B|A)P(A)}{P(B)}
\end{equation}

Where $A$ is our priors and $B$ is the posterior, i.e., observed data ($y$) I'm guessing? And so...

\begin{equation}
P(\mu,\sigma|y) = \frac{P(y|\mu,\sigma)P(\mu)P(\sigma)}{P(y)}
\end{equation}

Okay I really don't know...I guess we add in the model parameter definitions?

### E4 {.unnumbered}

> In the model definition below, which line is the linear model?
\begin{align}
y_i &\sim Normal(\mu,\sigma) \label{ex-e4a} \\
\mu &= \alpha + \beta x_i \label{ex-e4b} \\
\alpha &\sim Normal(0,10) \label{ex-e4c} \\
\beta &\sim Normal(0,1) \label{ex-e4d} \\
\sigma &\sim Exponential(2) \label{ex-e4e}
\end{align}

\ref{ex-e4b} is the linear model. \label{ex-e4a} is the likelihood, \label{ex-e4c} the prior for $\alpha$ (intercept), \label{ex-e4d} the prior for $\beta$ (slope), and \label{ex-e4e} the prior for sigma (sd).

### E5 {.unnumbered}

> In the model definition just above, how many parameters are in the posterior distribution?

The posterior distribution has the parameters:

- $\sigma$: standard deviation
- $\alpha$: intercept
- $\beta$: slope

### M1 {.unnumbered}

> For the model definition below, simulate observed y values from the prior (not the posterior).
\begin{align}
y_i &\sim Normal(\mu,\sigma) \label{ex-m1a} \\
\mu &\sim Normal(0,10) \label{ex-m1b} \\
\sigma &\sim Exponential(1) \label{ex-m1c}
\end{align}

```{r}
# same as R chunk 4.14
sample_mu <- rnorm(1e4, 0, 10)
sample_sigma <- rexp(1e4, 1)
# simulate prior predictive values
prior_h <- rnorm(1e4, sample_mu, sample_sigma)
dens(prior_h)
```


And our prior predictive with ggplot:

```{r}
n <- 1e4

set.seed(4)
tibble(sample_mu    = rnorm(n, mean = 0,       sd  = 10),
       sample_sigma = rexp(n, 1)) %>% 
  mutate(x = rnorm(n, mean = sample_mu, sd  = sample_sigma)) %>% 
  
  ggplot(aes(x = x)) +
  geom_density(fill = "red", linewidth = 0, alpha = .25) +
  scale_y_continuous(NULL, breaks = NULL) +
  labs(subtitle = expression(Prior~predictive~distribution~"for"~italic(h[i])),
       x = NULL) +
  theme(panel.grid = element_blank())
```

### Sampling from the posterior

```{r}
# R chunk 4.15
sample_mu <- rnorm( 1e4 , 178 , 100 ) 
prior_h <- rnorm( 1e4 , sample_mu , sample_sigma ) 
dens( prior_h )
```


```{r}
# R chunk 4.16
mu.list <- seq( from=150, to=160 , length.out=100 ) 
sigma.list <- seq( from=7 , to=9 , length.out=100 ) 
post <- expand.grid( mu=mu.list , sigma=sigma.list ) 
post$LL <- sapply( 1:nrow(post) , function(i) 
  sum( dnorm( d2$height , post$mu[i] , post$sigma[i] , log=TRUE ) ) ) 

post$prod <- post$LL + dnorm( post$mu , 178 , 20 , TRUE ) + dunif( post$sigma , 0 , 50 , TRUE ) 
post$prob <- exp( post$prod - max(post$prod) )
```

```{r}
# R chunk 4.17
sample.rows <- sample( 1:nrow(post) , size=1e4 , replace=TRUE , prob=post$prob ) 
sample.mu <- post$mu[ sample.rows ] 
sample.sigma <- post$sigma[ sample.rows ]
```

```{r}
# R chunk 4.21
dens(sample_mu)
```

```{r}
# R chunk 4.21
dens(sample_sigma)
```

### M2 {.unnumbered}

> Translate the model just above into a quap formula.

Do this as in R chunk 4.26:

```{r}
flist <-
  alist(height ~ dnorm(mu , sigma) , 
        mu ~ dnorm(0 , 10) , 
        sigma ~ dexp(0 , 1)
        )
```
